| Base de datos de abejas ibéricas \n| Iberian bees database\n

### 2022-10-06 Manuscript/manuscript.Rmd
Adding dot to rmd at the end of the paragraph

In the absence of standardized monitoring efforts, one may think that detailed information on species occurrences in space and time does not exist. However, there are different sources of extremely useful information. First, natural history museums hold historical collections of amateur naturalists and researchers who collected bees [@bartomeus2019]. Collectors are very good at classifying and labeling each collected specimen, which means we can retrieve the information stored along the pinned specimens about when and where those were collected. Second, the Iberian Peninsula has a rich tradition of pollinator and pollination researchers [@archer2014], which have been collecting occurrence bee data with different aims. Finally, many occurrence records are being made available through internet portals which centralize data hosted in institutions all around the world (e.g. Global Biodiversity Information Facility, GBIF) or collected by citizen scientists (e.g. iNaturalist). The challenge is how to digitalize, access, and curate all these different data sources. Thus, despite the obvious challenges to digitalize, access, and curate all these different data sources, they provide an essential source of information for biodiversity research [@nelson2019; @hedrick2020]

@@

Debido a la ausencia de esfuerzos de muestreo estandarizados se podría pensar que no existe información detallada del registro de especies en el espacio y en el tiempo. Sin embargo, hay diversas fuentes con información muy útil. En primer lugar, los museos de historia natural albergan colecciones históricas de naturalistas amateur e investigadores que capturaron abejas [@bartomeus2019]. Puesto que las colecciones de museos suelen estar muy bien clasificadas y etiquetadas, es posible recopilar información específica sobre la localización y momento de captura de los diferentes especímenes. En segundo lugar, la península ibérica tiene una rica tradición de investigadores sobre polinizadores y ecología de la polinización [@archer2014] que han recopilado gran cantidad de información de abejas con diferentes objetivos. Finalmente, muchos datos de ocurrencias han sido liberados de forma abierta a través de portales de internet que centralizan datos almacenados por instituciones de todo el mundo (p.ej. GBIF) o datos de ciencia ciudadana (p.ej. iNaturalist). Aunque sigue siendo un desafío digitalizar, acceder y depurar la información procedente de las diferentes fuentes de datos, la información que estas fuentes aportan ese esencial para la investigación de la biodiversidad [@nelson2019; @hedrick2020]

>>>>

In the absence of standardized monitoring efforts, one may think that detailed information on species occurrences in space and time does not exist. However, there are different sources of extremely useful information. First, natural history museums hold historical collections of amateur naturalists and researchers who collected bees [@bartomeus2019]. Collectors are very good at classifying and labeling each collected specimen, which means we can retrieve the information stored along the pinned specimens about when and where those were collected. Second, the Iberian Peninsula has a rich tradition of pollinator and pollination researchers [@archer2014], which have been collecting occurrence bee data with different aims. Finally, many occurrence records are being made available through internet portals which centralize data hosted in institutions all around the world (e.g. Global Biodiversity Information Facility, GBIF) or collected by citizen scientists (e.g. iNaturalist). The challenge is how to digitalize, access, and curate all these different data sources. Thus, despite the obvious challenges to digitalize, access, and curate all these different data sources, they provide an essential source of information for biodiversity research [@nelson2019; @hedrick2020].

@@

Debido a la ausencia de esfuerzos de muestreo estandarizados se podría pensar que no existe información detallada del registro de especies en el espacio y en el tiempo. Sin embargo, hay diversas fuentes con información muy útil. En primer lugar, los museos de historia natural albergan colecciones históricas de naturalistas amateur e investigadores que capturaron abejas [@bartomeus2019]. Puesto que las colecciones de museos suelen estar muy bien clasificadas y etiquetadas, es posible recopilar información específica sobre la localización y momento de captura de los diferentes especímenes. En segundo lugar, la península ibérica tiene una rica tradición de investigadores sobre polinizadores y ecología de la polinización [@archer2014] que han recopilado gran cantidad de información de abejas con diferentes objetivos. Finalmente, muchos datos de ocurrencias han sido liberados de forma abierta a través de portales de internet que centralizan datos almacenados por instituciones de todo el mundo (p.ej. GBIF) o datos de ciencia ciudadana (p.ej. iNaturalist). Aunque sigue siendo un desafío digitalizar, acceder y depurar la información procedente de las diferentes fuentes de datos, la información que estas fuentes aportan ese esencial para la investigación de la biodiversidad [@nelson2019; @hedrick2020].
